A Horrocks-type theorem for even orthogonal $K_2$

### 2020-01-26 dl-paper.tex
Minor language fixes

The proof of~\cref{thm:main} goes as follows. We notice that it suffices to prove the $\St(\rD_\ell, -)$-variant of the theorem. Moreover, the proof of the injectivity of
$j_+ \colon \St(\rD_\ell, A[X]) \to \St(\rD_\ell, A[X, X\inv])$ turns out to be the hardest part. After invoking the local-global principle~\cite[Theorem~2]{LS17} the proof reduces to the special case when $A$ is local.
Now if $M$ is the maximal ideal of $A$, the proof of the injectivity of $j_+$ comes down to proving injectivity of the following two maps:
\[ \xymatrix{ \St(\rD_\ell, A[X^{-1}]) \ar[r] & \St(\rD_\ell, A[X\inv] + M[X]) \ar[r] & \St(\rD_\ell, A[X, X\inv]). }\]
The injectivity of the second map is obtained in~\cref{thm41}. This is the only place in our proof which invokes the assumption that $2$ is invertible. The proof of~\cref{thm41} depends on the stability theorem for higher orthogonal K-groups and also on certain basic computations with Grothendieck--Witt groups (these groups include the stable orthogonal K-groups as a special case, see~\eqref{GW-concrete}). Finally, the injectivity of the first map is obtained in~\cref{thm:P1glueing}, which is a direct generalization of~\cite[Proposition~4.3]{Tu83}. This part of the proof is obtained in somewhat greater generality and is applicable to all simply-laced root systems $\Phi$ containing a subsystem of type $\rA_4$.
%$\rD_\ell$, $\ell=5,6$ and also $\rE_\ell$, $\ell=6,7,8$.

@@

In this paper we will be mostly interested in the case when the Dynkin diagram of $\Phi$ is {\it simply-laced}, i.\,e. does not contain double bonds. In this case the defining relations of $\St(\Phi, R)$ have the following simpler form:
\begin{align}
&x_{\alpha}(a)\cdot x_{\alpha}(b)=x_{\alpha}(a+b), \tag{R1}\\
&[x_{\alpha}(a),\,x_{\beta}(b)]=x_{\alpha+\beta}(N_{\alpha\beta} \cdot ab),\text{ for }\alpha+\beta\in\Phi, \tag{R2} \\
&[x_{\alpha}(a),\,x_{\beta}(b)]=1,\text{ for }\alpha+\beta\not\in\Phi\cup0. \tag{R3}
\end{align}
In the above formulae $a, b \in R$ and the integers $N_{\alpha, \beta} = N_{\alpha, \beta, 1, 1} = \pm 1$ are the structure constants of the Lie algebra of type $\Phi$. Although there is still some degree of freedom in their choice, they always must satisfy the relations, indicated in the following lemma (cf. \cite[\S~14]{VP}).
\begin{lemma} Suppose $\Phi$ is simply laced and $\alpha, \beta$ are roots of $\Phi$ such that $\alpha+\beta\in \Phi$, then holds
\begin{equation} \label{eq:simplest} N_{\alpha, \beta} = -N_{\beta,\alpha} = - N_{-\alpha, -\beta} = N_{\beta, -\alpha-\beta} = N_{-\alpha-\beta, \alpha}. \end{equation}
If, moreover, $\gamma \in \Phi$ is such that $\alpha,\beta,\gamma$ form a basis of a root subsystem of type $\rA_3$ then one has
\begin{equation} \label{eq:cocycle} N_{\beta,\gamma} \cdot N_{\alpha, \beta+\gamma} = N_{\alpha+\beta, \gamma} \cdot N_{\alpha, \beta}. \end{equation} \end{lemma}
In our computations below we will be using identities~\eqref{eq:simplest} without further reference.

@@

\subsection{\texorpdfstring{$\K_2$}{K2}-groups and symbols} In our computations we use two families of explicit elements of $\K_2(\Phi, R)$ called {\it Steinberg and Dennis--Stein symbols}. Notice that our notational conventions for symbols follow~\cite{DS73} and {\it not} more modern textbooks such as~\cite{Kbook}. Recall that Steinberg symbols are defined for arbitrary $s, t \in R^\times$ as follows:
\begin{equation} \label{eq:steinberg} \{ s, t \}_\alpha = h_\alpha(st) \cdot h_\alpha^{-1}(s) \cdot h_\alpha^{-1}(t). \end{equation}
In turn, Dennis--Stein symbols are defined for arbitrary $a, b\in R$ satisfying $1 + ab \in R^\times$:
\begin{equation} \label{eq:dennis-stein}  \langle a,b \rangle _ \alpha = x_{-\alpha}\left(\tfrac{- b}{1 + ab}\right) \cdot x_{\alpha}(a) \cdot x_{-\alpha}(b) \cdot x_{\alpha}\left(\tfrac{- a}{1+ab}\right) \cdot h_{\alpha}^{-1}(1 + ab). \end{equation} 
Dennis--Stein symbol $\langle a, b \rangle_\alpha$ can be expressed through Steinberg symbols in the special case when either $a$ or $b$ is an invertible element of $R$. More specifically, the following formulae hold (cf.~\cite[p.~250]{DS73}).
\begin{equation} \label{DS-S-relationship} \langle a, b \rangle_\alpha = \{-a, 1+ab\}_\alpha\text{ for } a, 1+ab\in R^\times,\ \
 \{ s, t \}_\alpha = \left\langle -s, \tfrac{1 - t}{s} \right\rangle_\alpha\text{ for } s, t\in R^\times. \end{equation}
Steinberg and Dennis--Stein symbols depend only on the length of $\alpha$, in particular they do not depend on $\alpha$ if $\Phi$ is simply-laced. If $\Phi$ happens to be {\it nonsymplectic}, i.\,e. $\Phi \neq \rA_1, \rB_2, \rC_{\geq 3}$, Steinberg symbols are antisymmetric and bimultiplicative, i.\,e. they satisfy the following identities: \begin{equation} \label{eq:symbol-properties} \{ u, st \} = \{ u, s\} \{ u, t \}, \ \{ u, v \} = \{ v, u\}^{-1}. \end{equation}
For these and other properties of symbols we refer the reader to~\cite{DS73}.

Recall that the classical Matsumoto theorem (see~\cite[Theorem~5.10]{Ma69}) allows one to compute the group $\K_2(\Phi, R)$ in the special case when $R=k$ is a field.
Using the modern language of Milnor--Witt K-theory (see~\cite{Mo04}) it can be formulated as the following computation:
\[\K_2(\Phi, k) = \left\{\begin{array}{ll} \K_2^\mathrm{MW}(k)& \text{if $\Phi$ is symplectic,}\\ \K_2^\mathrm{M}(k) & \text{otherwise.}\end{array}\right. \]
In the following lemma we recall the computation of the group $\K_2(\Phi, R)$ in the case $R=k[X, X\inv]$.
\begin{lemma}[Hurrelbrink--Morita--Rehmann]\label{K2-laurent-field} Let $\Phi$ be a reduced irreducible root system of type $\neq \rG_2$ and $k$ be arbitrary field. Then there is a split exact sequence of abelian groups
\[ \xymatrix{ 0 \ar[r] & \K_2(\Phi, k) \ar[r] & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & H(\Phi, k) \ar[r] & 0,}\text{ in which} \]
\[ H(\Phi, k) = \left\{\begin{array}{ll} \K_1^\mathrm{MW}(k)& \text{if $\Phi$ is symplectic,}\\ \K_1(k) \cong k^\times & \text{otherwise.}  \end{array}\right. \]  \end{lemma}
\begin{proof} Let us first consider the case of nonsymplectic $\Phi$, in which one can find a long root $\alpha\in \Phi$ in such a way that there is a commutative diagram of abelian groups
\[\xymatrixcolsep{5pc}\xymatrix{k^\times \ar[r]^-{h} \ar[rrd]_{\{-, X\}} & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & \K_2(\Phi, k(X)) \ar[d]^{\cong} \\
                                                                                      &                                 & \K_2^\mathrm{M}(k(X)),} \]
in which $h = \{ -, X \}_{\alpha}$ and the vertical map is an isomorphism by Matsumoto theorem.
Notice that the diagonal map is split by the obvious residue homomorphism and therefore is injective. This, in turn, implies that $h$ is also injective.
The assertion of the lemma now follows from~\cite[Satz~3]{Hur77} which asserts that for a nonsymplectic $\Phi$ holds $\K_2(\Phi, k[X, X\inv]) = \mathrm{Im}(h) \oplus \K_2(\Phi, k)$.

@@

\begin{externaltheorem} \label{thm41} Suppose that $2 \in A^\times$. Then for $\ell \geq 7$ the canonical map $\St(\rD_\ell, B) \to \St(\rD_\ell, R)$ is injective. \end{externaltheorem}
\begin{proof}
 Consider the following commutative diagram with exact rows, in which the lifting $t$ is obtained from~\cref{lem:lemma32}:
\begin{equation*} \xymatrix{
 C(\rD_\ell, B, I) \ar[r]^{\lambda_B} \ar@{->>}[d]_{f} & \St(\rD_\ell, B, I) \ar[r]^{\mu_B} \ar[d]_{g} &
 \St(\rD_\ell, B) \ar[r]^{\nu_B} \ar[d]_{h} & \St(\rD_\ell, k[X]) \ar[d]_{i} \\
 C(\rD_\ell, R, I) \ar[r]^{\lambda_R}         & \St(\rD_\ell, R, I) \ar[r]^{\mu_R} \ar@{-->}[ur]^{t}&
 \St(\rD_\ell, R) \ar[r]^-{\nu_R}        & \St(\rD_\ell, k[X, X^{-1}]).
}\end{equation*}
Let $a$ be an element of $\Ker(h)$. Since $i$ is injective by~\cref{field-injectivity}, the element $a$
 also lies in $\Ker(\nu_B)$ and hence comes from some $b \in \St(\rD_\ell, B, I)$ via $\mu_B$.
Since $g(b) \in \Ker(\mu_R)$ there exists some $c \in C(\rD_\ell, R, I)$ such that $\lambda_R(c) = g(b)$. 
By~\cref{lem:prop41}, $f$ is surjective, therefore $c = f(d)$ for some $d \in C(\rD_\ell, R, I)$.
The required assertion now follows from the following computation:
 \[ 1 = \mu_B\lambda_B(d) = tg\lambda_B(d) = t\lambda_Rf(d) =t(g(b)) = \mu_B(b) = a. \qedhere \]
\end{proof}

@@

The isomorphism between the sets $\overline{V}$ and $\overline{V}_T$ is given by the following two maps, which are easily seen to be mutually inverse to each other:
\begin{equation} \label{eq:VVT} \xymatrix @R0.3pc {\overline{V} \ar[r]^{\cong} & \ar[l] \overline{V}_T \\ (g, [h], u) \ar@{|->}[r] & [p(g, h, u), h^{-1}, u] \\ (p \cdot j_-(h) \cdot \{u, X\}, [h^{-1}], u) & \ar@{|->}[l] [p, h, u]. } \end{equation}
The above isomorphism allows us to regard $\overline{V}$ and $\overline{V}_T$ as one and the same object, for which we can interchangeably use either of the two notations,
 depending on which of them is more convenient in a given situation.
For example, specifying the action of $\St(\Phi, B)$ in terms of $\overline{V}$ leads to much shorter calculations in Lemmas~\ref{R3_leq0_1}--\ref{R2_0_1},
 while the statements of~\cref{prop43} and~\cref{lem:action} look more natural when formulated in terms of~$\overline{V}_T$.

>>>>

The proof of~\cref{thm:main} goes as follows. We notice that it suffices to prove the $\St(\rD_\ell, -)$-variant of the theorem. Moreover, the proof of the injectivity of
$j_+ \colon \St(\rD_\ell, A[X]) \to \St(\rD_\ell, A[X, X\inv])$ turns out to be the hardest part. After invoking the local-global principle~\cite[Theorem~2]{LS17} the proof reduces to the special case when $A$ is local.
Now if $M$ is the maximal ideal of $A$, the proof of the injectivity of $j_+$ comes down to proving injectivity of the following two maps:
\[ \xymatrix{ \St(\rD_\ell, A[X^{-1}]) \ar[r] & \St(\rD_\ell, A[X\inv] + M[X]) \ar[r] & \St(\rD_\ell, A[X, X\inv]). }\]
The injectivity of the second map is obtained in~\cref{thm41}. This is the only place in our proof which invokes the assumption that $2$ is invertible. The proof of~\cref{thm41} depends on the stability theorem for higher orthogonal K-groups and also on certain basic calculations with Grothendieck--Witt groups (these groups include the stable orthogonal K-groups as a special case, see~\eqref{GW-concrete}). Finally, the injectivity of the first map is obtained in~\cref{thm:P1glueing}, which is a direct generalization of~\cite[Proposition~4.3]{Tu83}. This part of the proof is obtained in somewhat greater generality and is applicable to all simply-laced root systems $\Phi$ containing a subsystem of type $\rA_4$.
%$\rD_\ell$, $\ell=5,6$ and also $\rE_\ell$, $\ell=6,7,8$.

@@

In this paper we will be mostly interested in the case when the Dynkin diagram of $\Phi$ is {\it simply-laced}, i.\,e. does not contain double bonds. In this case the defining relations of $\St(\Phi, R)$ have the following simpler form:
\begin{align}
&x_{\alpha}(a)\cdot x_{\alpha}(b)=x_{\alpha}(a+b), \tag{R1}\\
&[x_{\alpha}(a),\,x_{\beta}(b)]=x_{\alpha+\beta}(N_{\alpha\beta} \cdot ab),\text{ for }\alpha+\beta\in\Phi, \tag{R2} \\
&[x_{\alpha}(a),\,x_{\beta}(b)]=1,\text{ for }\alpha+\beta\not\in\Phi\cup0. \tag{R3}
\end{align}
In the above formulae $a, b \in R$ and the integers $N_{\alpha, \beta} = N_{\alpha, \beta, 1, 1} = \pm 1$ are the structure constants of the Lie algebra of type $\Phi$. Although there is still some degree of freedom in their choice, they always must satisfy the relations, indicated in the following lemma (cf. \cite[\S~14]{VP}).
\begin{lemma} Suppose $\Phi$ is simply laced and $\alpha, \beta$ are roots of $\Phi$ such that $\alpha+\beta\in \Phi$, then holds
\begin{equation} \label{eq:simplest} N_{\alpha, \beta} = -N_{\beta,\alpha} = - N_{-\alpha, -\beta} = N_{\beta, -\alpha-\beta} = N_{-\alpha-\beta, \alpha}. \end{equation}
If, moreover, $\gamma \in \Phi$ is such that $\alpha,\beta,\gamma$ form a basis of a root subsystem of type $\rA_3$ then one has
\begin{equation} \label{eq:cocycle} N_{\beta,\gamma} \cdot N_{\alpha, \beta+\gamma} = N_{\alpha+\beta, \gamma} \cdot N_{\alpha, \beta}. \end{equation} \end{lemma}
In our calculations below we will be using identities~\eqref{eq:simplest} without further reference.

@@

\subsection{\texorpdfstring{$\K_2$}{K2}-groups and symbols} In our calculations we use two families of explicit elements of $\K_2(\Phi, R)$ called {\it Steinberg and Dennis--Stein symbols}. Notice that our notational conventions for symbols follow~\cite{DS73} and {\it not} more modern textbooks such as~\cite{Kbook}. Recall that Steinberg symbols are defined for arbitrary $s, t \in R^\times$ as follows:
\begin{equation} \label{eq:steinberg} \{ s, t \}_\alpha = h_\alpha(st) \cdot h_\alpha^{-1}(s) \cdot h_\alpha^{-1}(t). \end{equation}
In turn, Dennis--Stein symbols are defined for arbitrary $a, b\in R$ satisfying $1 + ab \in R^\times$:
\begin{equation} \label{eq:dennis-stein}  \langle a,b \rangle _ \alpha = x_{-\alpha}\left(\tfrac{- b}{1 + ab}\right) \cdot x_{\alpha}(a) \cdot x_{-\alpha}(b) \cdot x_{\alpha}\left(\tfrac{- a}{1+ab}\right) \cdot h_{\alpha}^{-1}(1 + ab). \end{equation} 
Dennis--Stein symbol $\langle a, b \rangle_\alpha$ can be expressed through Steinberg symbols in the special case when either $a$ or $b$ is an invertible element of $R$. More specifically, the following formulae hold (cf.~\cite[p.~250]{DS73}).
\begin{equation} \label{DS-S-relationship} \langle a, b \rangle_\alpha = \{-a, 1+ab\}_\alpha\text{ for } a, 1+ab\in R^\times,\ \
 \{ s, t \}_\alpha = \left\langle -s, \tfrac{1 - t}{s} \right\rangle_\alpha\text{ for } s, t\in R^\times. \end{equation}
Steinberg and Dennis--Stein symbols depend only on the length of $\alpha$, in particular they do not depend on $\alpha$ if $\Phi$ is simply-laced. If $\Phi$ happens to be {\it nonsymplectic}, i.\,e. $\Phi \neq \rA_1, \rB_2, \rC_{\geq 3}$, Steinberg symbols are antisymmetric and bimultiplicative, i.\,e. they satisfy the following identities: \begin{equation} \label{eq:symbol-properties} \{ u, st \} = \{ u, s\} \{ u, t \}, \ \{ u, v \} = \{ v, u\}^{-1}. \end{equation}
For these and other properties of symbols we refer the reader to~\cite{DS73}.

Recall that the classical Matsumoto theorem (see~\cite[Theorem~5.10]{Ma69}) allows one to compute the group $\K_2(\Phi, R)$ in the special case when $R=k$ is a field.
Using the modern language of Milnor--Witt K-theory (see~\cite{Mo04}) it can be formulated as follows:
\[\K_2(\Phi, k) = \left\{\begin{array}{ll} \K_2^\mathrm{MW}(k)& \text{if $\Phi$ is symplectic,}\\ \K_2^\mathrm{M}(k) & \text{otherwise.}\end{array}\right. \]
In the following lemma we recall the computation of the group $\K_2(\Phi, R)$ in the case $R=k[X, X\inv]$.
\begin{lemma}[Hurrelbrink--Morita--Rehmann]\label{K2-laurent-field} Let $\Phi$ be a reduced irreducible root system of type $\neq \rG_2$ and $k$ be arbitrary field. Then there is a split exact sequence of abelian groups
\[ \xymatrix{ 0 \ar[r] & \K_2(\Phi, k) \ar[r] & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & H(\Phi, k) \ar[r] & 0,}\text{ in which} \]
\[ H(\Phi, k) = \left\{\begin{array}{ll} \K_1^\mathrm{MW}(k)& \text{if $\Phi$ is symplectic,}\\ \K_1(k) \cong k^\times & \text{otherwise.}  \end{array}\right. \]  \end{lemma}
\begin{proof} Let us first consider the case of nonsymplectic $\Phi$, in which one can find a long root $\alpha\in \Phi$ in such a way that there is a commutative diagram of abelian groups
\[\xymatrixcolsep{5pc}\xymatrix{k^\times \ar[r]^-{h} \ar[rrd]_{\{-, X\}} & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & \K_2(\Phi, k(X)) \ar[d]^{\cong} \\
                                                                                      &                                 & \K_2^\mathrm{M}(k(X)),} \]
in which $h = \{ -, X \}_{\alpha}$ and the vertical map is an isomorphism by Matsumoto theorem.
Notice that the diagonal map is split by the obvious residue homomorphism and therefore is injective. This, in turn, implies that $h$ is also injective.
The assertion of the lemma now follows from~\cite[Satz~3]{Hur77} which asserts that for a nonsymplectic $\Phi$ holds $\K_2(\Phi, k[X, X\inv]) = \mathrm{Im}(h) \oplus \K_2(\Phi, k)$.

@@

\begin{externaltheorem} \label{thm41} Suppose that $2 \in A^\times$. Then for $\ell \geq 7$ the canonical map $\St(\rD_\ell, B) \to \St(\rD_\ell, R)$ is injective. \end{externaltheorem}
\begin{proof}
 Consider the following commutative diagram with exact rows, in which the lifting $t$ is obtained from~\cref{lem:lemma32}:
\begin{equation*} \xymatrix{
 C(\rD_\ell, B, I) \ar[r]^{\lambda_B} \ar@{->>}[d]_{f} & \St(\rD_\ell, B, I) \ar[r]^{\mu_B} \ar[d]_{g} &
 \St(\rD_\ell, B) \ar[r]^{\nu_B} \ar[d]_{h} & \St(\rD_\ell, k[X]) \ar[d]_{i} \\
 C(\rD_\ell, R, I) \ar[r]^{\lambda_R}         & \St(\rD_\ell, R, I) \ar[r]^{\mu_R} \ar@{-->}[ur]^{t}&
 \St(\rD_\ell, R) \ar[r]^-{\nu_R}        & \St(\rD_\ell, k[X, X^{-1}]).
}\end{equation*}
Let $a$ be an element of $\Ker(h)$. Since $i$ is injective by~\cref{field-injectivity}, the element $a$
 also lies in $\Ker(\nu_B)$ and hence comes from some $b \in \St(\rD_\ell, B, I)$ via $\mu_B$.
Since $g(b) \in \Ker(\mu_R)$ there exists some $c \in C(\rD_\ell, R, I)$ such that $\lambda_R(c) = g(b)$. 
By~\cref{lem:prop41}, $f$ is surjective, therefore $c = f(d)$ for some $d \in C(\rD_\ell, R, I)$.
The required assertion now follows from the following calculation:
 \[ 1 = \mu_B\lambda_B(d) = tg\lambda_B(d) = t\lambda_Rf(d) =t(g(b)) = \mu_B(b) = a. \qedhere \]
\end{proof}

@@

The isomorphism between the sets $\overline{V}$ and $\overline{V}_T$ is given by the following two maps, which are easily seen to be mutually inverse to each other:
\begin{equation} \label{eq:VVT} \xymatrix @R0.3pc {\overline{V} \ar[r]^{\cong} & \ar[l] \overline{V}_T \\ (g, [h], u) \ar@{|->}[r] & [p(g, h, u), h^{-1}, u] \\ (p \cdot j_-(h) \cdot \{u, X\}, [h^{-1}], u) & \ar@{|->}[l] [p, h, u]. } \end{equation}
The above isomorphism allows us to regard $\overline{V}$ and $\overline{V}_T$ as the same object, for elements of which we can interchangeably use either of the two notations,
 depending on which of them is more convenient in a given situation.
For example, specifying the action of $\St(\Phi, B)$ in terms of $\overline{V}$ leads to much shorter calculations in Lemmas~\ref{R3_leq0_1}--\ref{R2_0_1},
 while the statements of~\cref{prop43} and~\cref{lem:action} look more natural when formulated in terms of~$\overline{V}_T$.
